\texttt{Taxonomic supertree construction with}\texttt{\emph{ Incertae sedis}}\texttt{ taxa}

### 2018-05-09 doc/incertae_sedis.tex
order of operations point

The rules for associated leaves in $\mathcal{I}(n)$ with state 1, leaves
     in $\mathcal{E}(n)$, and other leaves with the missing data symbol 
     do allow to encode the taxon $n$ in a taxonomic tree unambiguously.
However, these are not the only set of rules for generating a character matrix
    representation that would cause the taxonomic tree allows the characters to
    be mapped without homoplasy.
For example consider the case of the taxonomy shown in Figure \ref{fig:Placing-one-incertae}a.
Either of the two matrices shown in \ref{fig:Placing-one-incertae-MRP} could encode this
    taxonomy along with its \emph{incertae sedis} interpretation.

\begin{figure}
%\subfloat[Input tree \#1]{\includegraphics[scale=0.5]{Figures/is-within-is/tree1\lyxdot tre}}
\subfloat[Taxonomy]{\includegraphics[scale=0.5]{Figures/is-within-is/tax-shaded}
}
\subfloat[Synthesis tree]{\includegraphics[scale=0.5]{Figures/is-within-is/synth1-shaded}
}
\subfloat[Possible intermediate]{\includegraphics[scale=0.5]{Figures/is-within-is/intermed-shaded}
}

@@

\begin{figure}
%\subfloat[Input tree \#1]{\includegraphics[scale=0.5]{Figures/is-within-is/tree1\lyxdot tre}}
\subfloat[maximally scored]{
    \begin{tabular}{|r|p{2em}|p{2em}|}
\hline
& \multicolumn{2}{|c|}{Character} \\
taxa & $V$ & $WX$  \\
\hline
$U$  & 0 & 0  \\
$V1$ & 1 & ?  \\
$V2$ & 1 & ?  \\
$W$  & 0 & 1  \\
$X$  & 0 & 1 \\
\hline
\end{tabular}
}
\subfloat[less informative]{
    \begin{tabular}{|r|p{2em}|p{2em}|}
\hline
& \multicolumn{2}{|c|}{Character} \\
taxa & $V$ & $WX$  \\
\hline
$U$  & 0 & 0 \\
$V1$ & 1 & ? \\
$V2$ & 1 & ? \\
$W$  & ? & 1 \\
$X$  & 0 & 1 \\
\hline
\end{tabular}
}
\caption{\label{fig:Placing-one-incertae-MRP}
Two possible encodings that would support the $V$ and $WX$ branches and \emph{incertae sedis}
annotation shown in the taxonomy from in Figure \ref{fig:Placing-one-incertae}a.
The representations only vary in the scoring of leaf $W$ for character $V$.}
\end{figure}

>>>>

The rules for associated leaves in $\mathcal{I}(n)$ with state 1, leaves
     in $\mathcal{E}(n)$, and other leaves with the missing data symbol 
     do allow to encode the taxon $n$ in a taxonomic tree unambiguously.
However, these are not the only set of rules for generating a character matrix
    representation that would cause the taxonomic tree allows the characters to
    be mapped without homoplasy.
For example consider the case of the taxonomy shown in Figure \ref{fig:Placing-one-incertae}a.
Either of the two matrices shown in \ref{fig:Placing-one-incertae-MRP} could encode this
    taxonomy along with its \emph{incertae sedis} interpretation.
The matrix in Figure \ref{fig:Placing-one-incertae-MRP}a would result in the 
    taxon $V$ not being named in the supertree shown in Figure \ref{fig:Placing-one-incertae}b,
    while the matrix shown in Figure \ref{fig:Placing-one-incertae-MRP}b would lead
    to $V$ being named.
Thus, accurately
    honors the input taxonomic content we would need more taxonomic information
    than simply a hierarchy and knowledge of which taxa can float more tipward.

Note that this example also presents a challenge for iterative supertree building and refinement
    of the taxonomy without retention of additional information.
Consider the case, in which there are two phylogenetic inputs along with the taxonomy from
    tree Figure \ref{fig:Placing-one-incertae}b with the first phylogeny, $T_1$, providing the single
    rooted split: $V1\,X \mid \bullet\,U$ and the second phylogeny, $T_2$ providing the 
    rooted split: $W\, V1 \mid \bullet\,V2$.
If the supertree pipeline were run with just the taxonomy and and $T_1$ as inputs, then the supertree
    would be the tree shown in Figure \ref{fig:Placing-one-incertae}c.
One might hope that using that tree as the revised taxonomy and then adding the split from $T_2$ in
    a subsequent analysis would result in the same tree as if $T_1$ and $T_2$ to the original taxonomy.
While the topology for the final tree in either order of analyses would indeed be the tree shown in 
    Figure \ref{fig:Placing-one-incertae}b}, the decision about whether or not to name $V$ would differ.
Our handling of \emph{incertae sedis} in the original taxonomy via rooted partial splits
    would encode the taxonomic definitions in the manner depicted in Figure \ref{fig:Placing-one-incertae-MRP}a,
    but the encoding of the revised taxonomy in Figure \ref{fig:Placing-one-incertae}c would result
    in an altered definition of taxon $V$, as shown in \ref{fig:Placing-one-incertae-MRP}c.
As expected, in the encoding of the revised taxonomy, the definition of $WX$ has been altered to reflect the
    fact $T_1$ placed $V$ inside $WX$.
Perhaps, unexpected is the fact that storing the revised taxonomy with an \emph{incertae sedis} annotation
    and then reprocessing the taxonomy using our rooted partial splits approach would also shift
    the definition of $V$ from a taxon that must exclude $W$ to one that may include $W$.

\begin{figure}
%\subfloat[Input tree \#1]{\includegraphics[scale=0.5]{Figures/is-within-is/tree1\lyxdot tre}}
\subfloat[Taxonomy]{\includegraphics[scale=0.5]{Figures/is-within-is/tax-shaded}
}
\subfloat[Final supertree]{\includegraphics[scale=0.5]{Figures/is-within-is/synth1-shaded}
}
\subfloat[Possible intermediate]{\includegraphics[scale=0.5]{Figures/is-within-is/intermed-shaded}
}

@@

\begin{figure}
%\subfloat[Input tree \#1]{\includegraphics[scale=0.5]{Figures/is-within-is/tree1\lyxdot tre}}
\subfloat[maximally scored]{
    \begin{tabular}{|r|p{2em}|p{2em}|}
\hline
& \multicolumn{2}{|c|}{Character} \\
taxa & $V$ & $WX$  \\
\hline
$U$  & 0 & 0  \\
$V1$ & 1 & ?  \\
$V2$ & 1 & ?  \\
$W$  & 0 & 1  \\
$X$  & 0 & 1 \\
\hline
\end{tabular}
}
\subfloat[less informative]{
    \begin{tabular}{|r|p{2em}|p{2em}|}
\hline
& \multicolumn{2}{|c|}{Character} \\
taxa & $V$ & $WX$  \\
\hline
$U$  & 0 & 0 \\
$V1$ & 1 & ? \\
$V2$ & 1 & ? \\
$W$  & ? & 1 \\
$X$  & 0 & 1 \\
\hline
\end{tabular}
}
\subfloat[from revised taxonomy]{
    \begin{tabular}{|r|p{2em}|p{2em}|}
\hline
& \multicolumn{2}{|c|}{Character} \\
taxa & $V$ & $WX$  \\
\hline
$U$  & 0 & 0 \\
$V1$ & 1 & 1 \\
$V2$ & 1 & 1 \\
$W$  & ? & 1 \\
$X$  & 0 & 1 \\
\hline
\end{tabular}
}
\caption{\label{fig:Placing-one-incertae-MRP}
Panels (a) and (b) show two possible encodings that would support the $V$ and $WX$ branches and \emph{incertae sedis}
annotation shown in the taxonomy from in Figure \ref{fig:Placing-one-incertae}a.
The representations only vary in the scoring of leaf $W$ for character $V$. Panel (c) shows the information encoded by the tree in \ref{fig:Placing-one-incertae}c. }
\end{figure}
